---
title: "Microbial Biogeography of Restroom Surfaces"
author: "Thibaut Thalamas, Konstantinn Bonnet"
date: "12/15/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

Load packages.

```{r, results='hide'}
library(plyr)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(ggdendro)
library(RColorBrewer)
```

Set a general _ggplot2_ theme to use throughout.

```{r}
th <- theme_classic() +
	theme(panel.grid.major=element_line(size=0.5, linetype="solid", colour="#eeeeee"),
	panel.grid.minor=element_line(size=0.25, linetype="solid", colour="#eeeeee"))
theme_set(th)
```

## Read data and prepare data

Read the OTU table and cast counts to numeric,
since cor() and others require numeric data.

```{r}
# read otu table and cast counts to numeric, cor() and others require numeric data
otu <- read.delim("1335.vdb.tab", header=TRUE) %>%
	mutate_if(is.integer, as.numeric)
str(otu)
```

### Preliminary analysis and data manipulation

First, get the names of phyla and taxa by splitting the _OTUConTaxonomy_ column,
containing the assigned taxonomy for each OTU.

```{r}
# for each row of the otu table
tax <- apply(otu, 1, function(x)
	# select OTUConTaxonomy column
	x[2] %>%
		# split strings and get the result, returned inside a list
		strsplit(";") %>%
		# subset the list
		"[["(1) %>%
		# remove assignment score from the name
		sub(pattern="\\(.*", replacement="")
)
# transpose so OTUs are again the rows
tax <- t(tax)
str(tax)
# bind the new phylum and taxon column to the original table
otu <- cbind(otu, data.frame(Genus=tax[,2], Taxon=tax[,6]))
str(otu)
```

For analysis further on, extract the only the counts from the table,
and convert to a matrix: handy for some of the computations.

```{r}
otumat <- otu %>%
	# the counts are the only numeric columns
	select_if(is.numeric) %>%
	as.matrix
str(otumat)
```

Next, we'll need a the summed counts by taxon across OTUs,
and get the top 19 taxons in terms of abundance.
The very first element is actually _unclassified_,
the unassigned OTUs, and we'll discard it.

```{r}
# this same construct will be used later:
# take the numeric count matrix
taxtotals <- otumat %>%
	# for each of its columns (sites)
	apply(2, function(x)
		# for each unique taxon (rows)
		sapply(unique(otu$Taxon), function(t)
			# sum its abundance
			sum(x[otu$Taxon == t])
		)
	# this returns total abundance per taxon per site
	) %>%
	# sum abundance across sites => table with total abundance per taxon
	apply(1, sum)
str(taxtotals)
# order taxons by descending abundance
taxtotals <- taxtotals[order(-taxtotals)]
# then extract the top 19 taxons (removing unclassified)
toptax <- taxtotals[2:20]
toptax
```


## Phylum composition by site

We'll first expore phylum composition by site.
Using the previous manipulations,
we'll first compute a slightly prettier color palette for each phylum,
then output a stacked percent barchart,
showing relative proportion of each phylum in each site.

### Color palette

Here, we use ColorBrewer palettes for good constracts and for categorical data,
since those sets have the most colors.
We have more phylum (18) than colors in the set (12),
so we'll interpolate some additional colors.

```{r}
# number of unique phylum, the number of colors needed
ngenus <- length(unique(otu$Genus))
# extract the colors of the Set1 brewer color palette
# this gives a harmless warning since we request too many colors
pal.g <- brewer.pal(ngenus, "Set1")
# interpolate for the remaining elements
# colorRampPalette returns a function which we use directly to extract our colors
cmap.g <- colorRampPalette(pal.g, bias=1, interpolate="spline")(ngenus)
cmap.g
```

### Plot

We have to transform our OTU table into long format for use with _ggplot2_.
Since we don't reuse this result, we'll just pipe it directly to _ggplot()_.

Finally, we save the _ggplot_ object to a png file.
Rendering to pdf had different results on various machines,
as could possibly be seen in this document.

```{r, out.width='\\textwidth', fig.height=6, fig.align='center'}
# stacked percent barchart
bar.g <- otu %>%
	# get count columns and pivot to long format (similar to melt in reshape2)
	# tidyverse matches() is used to select the columns via a regex
	# site column will contains the sample site, and abundance the abundance values
	pivot_longer(cols=matches("^[a-z]", ignore.case=FALSE), names_to="site", values_to="abundance") %>%
	# x=site, y=abundance and use phylum as group colors for the bar's fill
	ggplot(aes(site, abundance, fill=Genus)) +
		# output percent stacked barchart
		geom_bar(position="fill", stat="identity") +
		# rotate xlab and remove x title
		theme(axis.text.x=element_text(angle=45, vjust=1, size=12, hjust=1),
			axis.title=element_blank()) +
		# manually set the colors for each phylum from the palette
		scale_fill_manual(values=cmap.g[order(unique(otu$Genus))])
bar.g
# save it
ggsave("plots/stacked.genus.by.site.png", bar.g, width=12, height=10)
```

## Taxonomic composition by site

Next, we'll do the same, but this time for taxa.
These are more precise and familiar and can provide interesting results.

To make a prettier and more informative plot,
there are several complicated steps involved.

To easily view which phylum the taxa belong to,
we'd like to group them visually.
An intuitive way to do this is to make gradients of one color for each phylum,
since there are no more than a few of them,
and the taxa are few enough that there still will be enough contrast to distinguish each of them.

To this end, we need to create the color palette manually for each,
but in a scriptable and extensible way.

### Creating the color palettes

We'll have to build a correspondence table
with the top 19 taxa found earlier.

First, we extract the phylum and taxon columns from the taxonomy table computed earlier.

```{r}
# filter the taxonomy table by the names of the top taxa, get phylum (2) and taxon (6) columns
corr <- tax[tax[,6] %in% names(toptax),][,c(2,6)]
# only take unique rows, then convert to data frame and set appropriate column names
corr <- unique(corr) %>%
	as.data.frame %>%
	# "family" seems more appropriate than "genus" here
	rename(Family=V1, Taxon=V2)
corr
```

Next, we use ordered factors to force ordering by phylum, then taxon.
This is necessary to preserve this order later since we further manipulate the tables,
and have to make sure colors are correctly attributed.

```{r}
# take our correspondence table
corr <- corr %>%
	# convert the family column to an ordered factor according to abundance and importance
	# the levels are from what was obtained above
	mutate(Family=factor(Family,
		levels=c("Proteobacteria", "Actinobacteria", "Firmicutes", "Bacteroidetes"),
		ordered=TRUE)) %>%
	# sort by family then taxon
	arrange(Family, Taxon) %>%
	# now freeze the new taxon order
	mutate(Taxon=factor(Taxon, levels=unique(Taxon), ordered=TRUE))
corr
```

Finally, we can attribute color gradients for each phylum,
making sure each gets a different base color.

```{r}
# add a new column with the colors for the ordering
corr$col <- c(
	# blue, red, green and violet/pink gradients
	# like before, we directly call the function returned with the appropriate number of colors we need
	colorRampPalette(c("lightblue", "darkblue"))(8),
	colorRampPalette(c("#ffff66", "red", "darkred"))(5),
	colorRampPalette(c("lightgreen", "darkgreen"))(4),
	colorRampPalette(c("violet", "blue"))(8)[1:2]
)
corr
```

### Stacked percent barchart

We can now output the final plot using the colors we computed,
the same way as before.

```{r, out.width='\\textwidth', fig.height=6, fig.align='center'}
bar.t <- otu %>%
	# remove columns we don't care about
	select(-Genus, -starts_with("OTU")) %>%
	# extract rows for only top taxa
	filter(Taxon %in% names(toptax)) %>%
	# convert to long format for ggplot as before
	pivot_longer(cols=matches("^[a-t]", ignore.case=FALSE), names_to="site", values_to="abundance") %>%
	# convert to ordered factor as with the correspondence table, using its order
	# counterintuitive and timeconsuming to figure out, but very important to avoid any discrepancies...
	mutate(Taxon=factor(Taxon, levels=levels(corr$Taxon), ordered=TRUE)) %>%
	ggplot(aes(site, abundance, fill=Taxon)) +
		# stacked percent barchart
		geom_bar(position="fill", stat="identity") +
		# set our colors
		scale_fill_manual(values=corr$col) +
		# rotate x legend, remove x axis label
		theme(axis.text.x=element_text(angle=45, vjust=1, size=12, hjust=1),
			axis.title=element_blank())
bar.t
# save the plot
ggsave("plots/stacked.taxon.by.site.png", bar.t, width=12, height=10)
```

## Phylum correlation map

It is of interest to evaluate links between phylum in our data.
This can be done by computing the correlation between phylum,
then visualizing it as a sort of heatmap.

### Preparing the data

We first need to extract the totals for each phylum by site,
since that's what we'll be calculating the correlation matrix on.

```{r}
# use the abundance matrix computed at the beginning, cor() needs numerics
gentotals <- otumat %>%
	# for each site (column)
        apply(2, function(x)
		# for each OTU (row),
		# sum abundances of OTUs assigned to each individual phylum
                sapply(unique(otu$Genus), function(g)
                        sum(x[otu$Genus == g])
                )
        )
head(gentotals)
```

There is a high number of OTUs with single counts across sites,
and many unassigned ones.
Similarly, there are phylums with only a few corresponding rows,
and very low counts.
We could still compute a correlation with the rest,
but it would be difficult to interpret.
We set the threshold as low as possible.
Two phylum are removed.

```{r}
# calculate a frequency table 
genfreq <- table(otu$Genus)
genfreq
# filter the most infrequent phylum
gentotals <- gentotals[!rownames(gentotals) %in% names(genfreq)[genfreq < 5],]
head(gentotals)
```

### Computing correlation and ordering

We'll first compute an ordinary correlation matrix between phylum,
which are the rows of the total abundance matrix.

```{r}
gencor <- cor(t(gentotals))
```

Next, we'd like to group highly correlated phylum together.
An easy way is to perform hierarchical clustering and use the clustering's order.
We even get a dendrogram for easier visualization that way.

For this kind of data, Pearson distance, defined as 1 - Pearson correlation is appropriate,
and Ward linkage method works very well.
As indicated in the manual, _ward.D2_ must be used, since our distances are not squared (or positive integers).

```{r}
# perform the clustering
hcl <- hclust(as.dist(1 - gencor), method="ward.D2")
# extract order
horder <- rownames(gentotals)[hcl$order]
horder
# convert to data appropriate for ggplot2, which gives us segment coordinates for drawing it directly on top
dhcl <- dendro_data(hcl)
# reorder the matrix
gencor <- gencor[horder,horder]
gencor
```

We then remove redundant data by removing the top triangle of the matrix.

```{r}
# lower.tri selects the matrix' lower triangle, and we just remove all the data
gencor[lower.tri(gencor)] <- NA
```

## Correlation map

We can now output a correlation map similar similar to a heatmap.
Colors are set as white for no correlation and red and blue for positive and negative correlation respectively.
We add the dendrogram obtained before above it.

```{r, out.width='\\textwidth', fig.height=6, fig.align='center'}
# use the ordered correlation matrix
cormap <- gencor %>%
	# convert to dataframe for ggplot
	as.data.frame %>%
	# get the rownames to use for grouping
	rownames_to_column("grp") %>%
	# convert to long format, remove missing values (upper triangle)
	pivot_longer(-grp, names_to="key", values_to="v", values_drop_na=TRUE) %>%
	# make sure the order is preserved like before, otherwise elements may be shifted seemingly at random
	mutate(grp=factor(grp, levels=horder, ordered=TRUE),
		key=factor(key, levels=horder, ordered=TRUE)) %>%
	# pass it directly to ggplot; label will be the rounded correlation value within each cell
	ggplot(aes(key, grp, fill=v, label=round(v,2))) +
			# tilemap background
			geom_tile(color="white") +
			# set a blue to red gradient for [-1;1] correlation values
			scale_fill_gradient2(
				low="blue", high="red", mid="white",
				midpoint=0,
				limit=c(-1,1), space="Lab",
				name="Pearson\nCorrelation") +
			# rotate x axis for easier viewing, remove axis title
			theme(axis.text.x=element_text(angle=45, vjust=1, size=12, hjust=1),
				axis.title=element_blank()) +
			# make sure we get a square matrix
			coord_fixed() +
			# add correlation values
			geom_text(color="black", size=3.5) +
			# draw the dendrogram; we offset y coordinates to place it above the tilemap
			geom_segment(data=segment(dhcl), aes(x, y+nrow(gencor)+0.5, xend=xend, yend=yend+nrow(gencor)+0.5), inherit.aes=FALSE)
cormap
# save it
ggsave("plots/cormap.pdf", cormap, width=12, height=10)
```


## Principal component analysis

Finally, we can perform a simple PCA to evaluate if there is any grouping structure in the data.

```{r}
# compute pca on numeric matrix, scale the variables to unit variance as recommended
pca <- prcomp(otumat, scale=TRUE)
# percent of variance explained by each axis
pca$sdev ^ 2 / sum(pca$sdev ^ 2)
# 40.6% and 28.6% for PC1 and PC2
```

```{r, out.width='\\textwidth', fig.height=6, fig.align='center'}
# render base graphics plot
biplot(pca,
	# draw arrows for each eigenvector
	var.axes=TRUE,
	xlabs=rep("*", nrow(otumat)), cex=c(0.9,1.1),
	xlim=c(-0.9,0.2), expand=50,
	col=c("black", "#dd0000"))
```

```{r, results='hide'}
pdf("plots/pca.pdf", 12, 10)
# plot the pca
biplot(pca,
	# draw arrows for each eigenvector
	var.axes=TRUE,
	xlabs=rep("*", nrow(otumat)), cex=c(0.9,1.1),
	xlim=c(-0.9,0.2), expand=50,
	col=c("black", "#dd0000"))
dev.off()
```
